### 2018-07-16 latex/results.tex
Incorporated changes by Toni

To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPRT, NADA and NNMT), we first conducted a comprehensive phylogenetic analysis. The phylogenetic distribution of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT is scattered in bacteria \cite{Gazzaniga2009}, while their co-occurrence has been detected in some marine invertebrates \cite{Gossmann2012FEBS}. As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT. In contrast, Metazoa predominantly lost NADA and have NamPRT together with NNMT. NNMT seems to have arisen \textit{de~novo} in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B). We were unable to find any \textit{NNMT} gene with an e-value below 0.1 in fungi or plants.

Nematodes are the only organisms, where we observed a concomitant presence of NADA and NNMT. In deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT are Sauropsida, and among them especially birds. The reason why about half of the sequenced bird genomes do not seem to encode for NNMT remains unclear. The distribution of NNMT in birds is quite scattered (suppl. fig.~S2). The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine. There are few metazoan species for which we could not find NamPRT or NADA, while NNMT was present. We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.

In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions. To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \cite{Gossmann2012FEBS}. The detailed list of templates used for the phylogenetic analyses can be found in supplementary table~S1. The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each clade. With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in protostomes, whereas most deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.

@@

To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we had limited information about expression levels of enzymes, we initially used equal amounts for all enzymes. As we have very few cross species kinetic data, we were furthermore mainly relying on kinetic constants found for human or yeast enzymes. Wherever possible, we included both substrate affinities and known product inhibitions or inhibition by downstream metabolites. As we in addition assumed that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.

First, we addressed the unexpected correlation between the selection for co-occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families. We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT. To achieve free NAD concentrations in the range reported in the literature and due to the very low turnover of NamPRT, we used tenfold higher NamPRT levels compared to other enzymes. We also adjusted the amount of NMNAT accordingly to avoid that the NAD synthesis rates are limited by this enzyme. Surprisingly, as shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher rather than lower NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A). However, it diminishes the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B). The decline in NAD concentration can be compensated by a higher expression of NamPRT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).

@@

Kinetic paramters of NamPRT were previously meassured for the human enzyme \cite{Burgos2008} as well as for some bacterial enzymes \cite{Sorci2010}, the latter having a much lower substrate affinity for Nam. We thus analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux. In the absence of NNMT, a variation of the substrate affinity of NamPRT for Nam, has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B). In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values of NamPRT (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).

@@

Given that NNMT might have exerted an evolutionary pressure on the development of NamPRT, one would expect to observe adaptations that are reflected in the NamPRT protein sequence arising with the occurrence of NNMT. To explore this, we created a multiple sequence alignment. The alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1. We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme. This insert overlaps with a predicted weak nuclear localisation signal (NLS), that is lost when the insert is removed. These ten amino acids correspond to a stretch at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g. structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}). Intriguingly, this presumed loop, depicted in red in figure~\ref{fig:unresolved_loop}B, is connected to one of the $\beta$-sheets involved in substrate binding \cite{Burgos2009} and in the functional homodimer, the two loops are placed side-by-side.

@@

The second hypothesis was that the sequence insertion might influence substrate binding of NamPRT. We thus expressed both wildtype and mutant proteins, N-terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them. The size exclusion chromatography profile showed that both wildtype and mutant protein were expressed as dimers (see suppl. fig.~S3), indicating that the missing residues in the mutant did not dramatically affect the protein structure. The enzymatic activity was measured by NMR spectroscopy using the detection of NMN produced in the presence and absence of ATP. Upon incubation with the NamPRT inhibitor FK866 \cite{Hasmann2003} for 30 minutes, both wildtype and mutant NamPRT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl. fig.~S4). Nevertheless, the enzymatic activity of the mutant enzyme was only approximately 30\% of that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D). In contrast to wildtype NamPRT, the activity of the mutant enzyme was not stimulated in the presence of ATP (fig.~\ref{fig:unresolved_loop}E). These observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866. However, the lower activity and the absence of catalytic activation by ATP indicate that deletion of amino acids 42 to 51 might have affected substrate binding.

>>>>

To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPRT, NADA and NNMT), we first conducted a comprehensive phylogenetic analysis of these three enzymes. The phylogenetic distribution of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT is scattered in bacteria \citep{Gazzaniga2009}, while their co\-/occurrence has been detected in some marine invertebrates \citep{Gossmann2012FEBS}. As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT. In contrast, Metazoa predominantly lost NADA and have NamPRT together with NNMT. NNMT seems to have arisen \textit{de~novo} or diverged rapidly in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B). We were unable to find any \textit{NNMT} gene with an e-value below 0.1 in fungi or plants.

Nematodes are the only organisms, where we observed a concomitant presence of NADA and NNMT. In deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT are Sauropsida, and among them especially birds. The reason why about half of the sequenced bird genomes do not seem to encode for NNMT remains unclear. The distribution of NNMT in birds is quite scattered (suppl. fig.~S2) but could be explained by the fact that numerous bird genes are high in GC content \citep{Hron2015}. The lack of NNMT might be related to the differences in the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine. There are few metazoan species for which we could not find NamPRT or NADA, while NNMT was present. We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.

In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions. To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \citep{Gossmann2012FEBS}. The detailed list of templates used for the phylogenetic analyses can be found in supplementary table~S1. The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each clade. With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in protostomes, whereas most deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \citep{Gossmann2014DNAR}.

@@

To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we had limited information about species-specific expression levels of enzymes, we initially assumed equal expression rates for all enzymes. As we have very few cross species kinetic data, we were furthermore mainly relying on kinetic constants found for human or yeast enzymes. Wherever possible, we included both substrate affinities and known product inhibitions or inhibition by downstream metabolites. As we in addition assumed that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.

First, we addressed the unexpected correlation between the selection for co\-/occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families. We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT. To achieve free NAD concentrations in the range reported in the literature and due to the very low turnover of NamPRT, we used tenfold higher NamPRT levels compared to other enzymes. We also adjusted the amount of NMNAT accordingly to avoid that the NAD synthesis rates are limited by this enzyme. Surprisingly, as shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher rather than lower NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A). However, it diminishes the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B). The decline in NAD concentration can be compensated by a higher expression of NamPRT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).

@@

Kinetic paramters of NamPRT were previously meassured for the human enzyme \citep{Burgos2008} as well as for some bacterial enzymes \citep{Sorci2010}, the latter having a much lower substrate affinity for Nam. We thus analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux. In the absence of NNMT, a variation of the substrate affinity of NamPRT for Nam, has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B). In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values of NamPRT (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).

@@

Given that NNMT might have exerted an evolutionary pressure on the development of NamPRT, one would expect to observe adaptations that are reflected in the NamPRT protein sequence arising shortly after the occurrence of NNMT. To explore this, we created a multiple sequence alignment. The alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1. We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme. This insert overlaps with a predicted weak nuclear localisation signal (NLS), that is lost when the insert is removed. These ten amino acids correspond to a stretch at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g. structure visualisation in fig.~\ref{fig:unresolved_loop}B from \citet{Wang2006}). Intriguingly, this presumed loop, depicted in red in figure~\ref{fig:unresolved_loop}B, is connected to one of the $\beta$-sheets involved in substrate binding \citep{Burgos2009} and in the functional homodimer, the two loops are placed side-by-side.

@@

The second hypothesis was that the sequence insertion might influence substrate binding of NamPRT. We thus expressed both wildtype and mutant proteins, N\=/terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them. The size exclusion chromatography profile showed that both wildtype and mutant protein were expressed as dimers (see suppl. fig.~S3), indicating that the missing residues in the mutant did not dramatically affect the protein structure. The enzymatic activity was measured by NMR spectroscopy using the detection of NMN produced in the presence and absence of ATP. Upon incubation with the NamPRT inhibitor FK866 \citep{Hasmann2003} for 30 minutes, both wildtype and mutant NamPRT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl. fig.~S4). Nevertheless, the enzymatic activity of the mutant enzyme was only approximately 30\% of that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D). In contrast to wildtype NamPRT, the activity of the mutant enzyme was not stimulated in the presence of ATP (fig.~\ref{fig:unresolved_loop}E). These observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866. However, the lower activity and the absence of catalytic activation by ATP indicate that deletion of amino acids 42 to 51 might have affected substrate binding.
